# Search & Regular Expression {#command-line-regular-expression}

In this chapter, we will explore commands that will

- search for a given string in a file
- find files using names
- and search for binary executable files

```{r table_search, echo=FALSE}
cname   <- c("`grep`", "`find`", "`which`")
descrip <- c("Search for a given string in a file",
             "Find files using filenames",
             "Search for binary executable files")
data.frame(Command = cname, Description = descrip) %>% 
  kable() %>% 
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "responsive")
  )
```

## grep

The `grep` command is used for pattern matching. Along with additional options, it can be used to

- match pattern in input text
- ignore case
- search recursively for an exact string
- print filename and line number for each match
- invert match for excluding specific strings

`grep` processes text line by line, and prints any lines which match a specified pattern. `grep`, which stands for **g**lobal **r**egular **e**xpression **p**rint is a powerful tool for matching a regular expression against text in a file, multiple files, or a stream of input. 

```{r table_grep, echo=FALSE}
cname   <- c("`grep`", "`grep -i`", "`grep -RI`", "`grep -E`", "`grep -Hn`", "`grep -v`")
descrip <- c("Matches pattern in input text",
             "Ignore case",
             "Search recursively for an exact string",
             "Use extended regular expression",
             "Print file name & corresponding line number for each match",
             "Invert match for excluding specific strings")
data.frame(Command = cname, Description = descrip) %>% 
  kable() %>% 
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "responsive")
  )
```

### Match Pattern in Input Text

Using `grep` let us search for packages that inlcude the letter `R` in their names.

```{bash c57a, eval=FALSE}
grep R package_names.txt
```

```{bash c57b, echo=FALSE}
cd cline
grep R package_names.txt
```

### Ignore Case

In the previous case, `grep` returned only those packages whose name included `R` but not `r` i.e. it did not ignore the case of the letter. Using the `-i` option, we will now search while ignoring the case of the letter.

```{bash c58a, eval=FALSE}
grep -i R package_names.txt
```

```{bash c58b, echo=FALSE}
cd cline
grep -i R package_names.txt
```

### Highlight

The `--color` option will highlight the matched strings. 

```{bash c85a, eval=FALSE}
grep -i --color R package_names.txt
```

```{bash c85b, echo=FALSE}
cd cline
grep -i --color R package_names.txt
```

### Print Filename 

If there is more than one file to search, use the `-H` option to print the filename for each match. 

```{bash c86a, eval=FALSE}
grep -i --color -H bio package_names.txt
```

```{bash c86b, echo=FALSE}
cd cline
grep -i --color -H bio package_names.txt
```

### Print Corresponding Line Number

The `-n` option will print the corresponding line number of the match in the file.

```{bash c87a, eval=FALSE}
grep -i --color -n bio package_names.txt
```

```{bash c87b, echo=FALSE}
cd cline
grep -i --color -n bio package_names.txt
```

### Print Filename & Line Number

Let us print both the file name and the line number for each match.

```{bash c59a, eval=FALSE}
grep -i --color -Hn R package_names.txt
```

```{bash c59b, echo=FALSE}
cd cline
grep -i --color -Hn R package_names.txt
```

### Invert Match

Use the `-v` option to select non-matching lines. In the below example, we search for packages whose name does not include `R` while ignoring the case.

```{bash c60a, eval=FALSE}
grep -v -i R package_names.txt
```

```{bash c60b, echo=FALSE}
cd cline
grep -v -i R package_names.txt
```

### Recursive Search

Use the `-r` option to search recursively. In the below example, we search all files with the `.txt` extension for the string `bio` while ignoring the case.

```{bash c88a, eval=FALSE}
grep -i --color -r bio *.txt
```

```{bash c88b, echo=FALSE}
cd cline
grep -i --color -r bio *.txt
```

## find

The `find` command can be used for searching files and directories. Using additional options, we can

- search files by extension type
- ignore case while searching files/directories

`find` is a powerful tool for working with the files. It can be used on its own to locate files, or in conjunction with other programs to perform operations on those files.

```{r table_find, echo=FALSE}
cname   <- c("`find`", "`find -name '*.txt'`", "`find -type d -iname`", "`find -type d -name`")
descrip <- c("Find files or directories under the given directory; recursively",
             "Find files by extension",
             "Find directories matching a given name, in case-insensitive mode",
             "Find directories matching a given name, in case-sensitive mode")
data.frame(Command = cname, Description = descrip) %>% 
  kable() %>% 
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "responsive")
  )
```

### Search Recursively

Let us use `find` to search for the file `release_names.txt` recursively. The `-name` option is used to specify the name of the file we are searching.

```{bash c61a, eval=FALSE}
find -name release_names.txt
```

```{bash c61b, echo=FALSE}
cd cline
find -name release_names.txt
```

There are two files with the name `release_names.txt` present in the current working directory and in `r_releases` directory.

### Search by Extension

Let us search for all files with `.txt` extension in the `r_releases` folder.

```{bash c62a, eval=FALSE}
find r_releases -name '*.txt'
```

```{bash c62b, echo=FALSE}
cd cline
find r_releases -name '*.txt'
```

There are 3 files with the `.txt` extension in `r_releases` folder.

### Case-insensitive Mode

Search for all folders with the name `R` or `r`. Here we use the `-iname` option to ignore case while searching. The `-type` option is used to specify whether we are searching for files or folders. Since we are searching for folder/directory, we use it along with `d` i.e. directory to indicate that we are searching for directories and not files.

```{bash c63a, eval=FALSE}
find -type d -iname R
```

```{bash c63b, echo=FALSE}
cd cline
find -type d -iname R
```

### Case-sensitive Mode

Search for all folders with the name `r`. It should exclude any folder with the name `R`.

```{bash c64a, eval=FALSE}
find -type d -name r
```

```{bash c64b, echo=FALSE}
cd cline
find -type d -name r
```