# Data Transfer {#datatransfer}

In this chapter, we will explore commands that will allow us to download files 
from the internet.

```{r table_data_transfer, echo=FALSE}
cname   <- c("`wget`", "`curl`", "`hostname`", "`ping`", "`nslookup`")
descrip <- c("Download files from the web",
             "Transfer data from or to a server",
             "Name of the current host",
             "Ping a remote host",
             "Name server details")
data.frame(Command = cname, Description = descrip) %>% 
  kable() %>% 
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "responsive")
  )
```

We have not executed the commands in this ebook as downloading multiple files from the internet will take a lot of time or result in errors but we have checked all the commands offline to ensure that they work.

## wget

The `wget` command will download contents of a URL and files from the internet. 
Using additional options, we can

- download contents/files to a file
- continue incomplete downloads
- download multiple files 
- limit download speed and number of retries

```{r table_wget, echo=FALSE}
cname   <- c("`wget url`", "`wget -o file url`", "`wget -c`", "`wget -P folder_name -i urls.txt`", "`wget --limit-rate`", "`wget --tries`",
              "`wget --quiet`", "`wget --no-verbose`", "`wget --progress-dot`", "`wget --timestamping`", "`wget --wait`")
descrip <- c("Download contents of a url",
             "Download contents of url to a file",
             "Continue an incomplete download",
             "Download all urls stored in a text file to a specific directory",
             "Limit download speed", 
             "Limit number of retries",
             "Turn off output",
             "Print basic information",
             "Change progress bar type to dot",
             "Check if the timestamp of the file has changed before downloading",
             "Wait between retrievals")
data.frame(Command = cname, Description = descrip) %>% 
  kable() %>% 
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "responsive")
  )
```

### Download URL

Let us first use `wget` to download contents of a URL. Note, we are not downloading file as such but just the content of the URL. We will use the URL of the home page of [R](https://www.r-project.org/) project. 

```{bash c113a, eval=FALSE}
wget https://www.r-project.org/
```

```{bash c113b, echo=FALSE, eval=FALSE}
cd cline
wget https://www.r-project.org/
ls
```

If you look at the list of files, you can see a new file, `index.html` which we just downloaded using `wget`. Downloading contents this way will lead to confusion if we are dealing with multiple URLs. Let us learn to save the contents to a file (we can specify the name of the file which should help avoid confusion.)

### Specify Filename

In this example, we download contents from the same URL and in addition specify the name of the file in which the content must be saved. Here we save it in a new file, `rhomepage.html` using the `-o` option followed by the filename.

```{bash c114a, eval=FALSE}
wget -o rhomepage.html https://www.r-project.org/
```

```{bash c114b, echo=FALSE, eval=FALSE}
cd cline
wget -o rhomepage.html https://www.r-project.org/
ls
```

### Download File

How about downloading a file instead of a URL? In this example, we will download a logfile from the [RStudio CRAN](http://cran-logs.rstudio.com/) mirror. It contains the details of R downloads and individual package downloads. If you are a package developer and would want to know the countries in which your packages are downloaded, you will find this useful. We will download the file for 29th September and save it as `sep_29.csv.gz`.

```{bash c108a, eval=FALSE}
wget -o sep_29.csv.gz http://cran-logs.rstudio.com/2019/2019-09-29.csv.gz
```

```{bash c108b, echo=FALSE, eval=FALSE}
cd cline
wget -o sep_29.csv.gz http://cran-logs.rstudio.com/2019/2019-09-29.csv.gz
```

### Download Multiple URLs 

How do we download multiple URLs? One way is to specify the URLs one after the other separated by a space or save all URLs in a file and read them one by one. In the below example, we have saved multiple URLs in the file `urls.txt`. 

```{bash c115a, eval=FALSE}
cat urls.txt
```

```{bash c115b, echo=FALSE}
cd cline
cat urls.txt
```

We will download all the above URLs and save them in a new folder `downloads`. The `-i` indicates that the URLs must be read from a file (local or external). The `-P` option allows us to specify the directory into which all the files will be downloaded. 

```{bash c109a, eval=FALSE}
wget -P downloads -i urls.txt     
```

### Quiet

The `--quiet` option will turn off `wget` output. It will not show any of the following details:

- name of the file being saved
- file size
- download speed
- eta etc.

```{bash c110a, eval=FALSE}
wget –-quiet http://cran-logs.rstudio.com/2019/2019-10-06.csv.gz
```

### No Verbose

Using the `-nv` or `--no-verbose` option, we can turn off verbose without being completely quiet (as we did in the previous example). Any error messages and basic information will still be printed.

```{bash c111a, eval=FALSE}
wget –-no-verbose http://cran-logs.rstudio.com/2019/2019-10-13.csv.gz    
```

### Check Timestamp

Let us say we have already downloaded a file from a URL. The file is updated from time to time and we intend to keep the local copy updated as well. Using the `--timestamping` option, the local file will have timestamp matching the remote file; if the remote file is not newer (not updated), no download will occur i.e. if the timestamp of the remote file has not changed it will not be downloaded. This is very useful in case of large files where you do not want to download them unless they have been updated.

```{bash c112a, eval=FALSE}
wget –-timestamping http://cran-logs.rstudio.com/2019/2019-10-13.csv.gz
```

## curl

The `curl` command will transfer data from or to a server. We will only look at
downloading files from the internet.

```{r table_curl, echo=FALSE}
cname   <- c("`curl url`", "`curl url -o file`", "`curl url > file`", "`curl -s`")
descrip <- c("Download contents of a url",
             "Download contents of url to a file",
             "Download contents of url to a file",
             "Download in silent or quiet mode")
data.frame(Command = cname, Description = descrip) %>% 
  kable() %>% 
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "responsive")
  )
```

### Download URL

Let us download the home page of the [R](https://www.r-project.org/) project using `curl`.

```{bash c65a, eval=FALSE}
curl https://www.r-project.org/
```

```{bash c65b, echo=FALSE, eval=FALSE}
cd cline
curl https://www.r-project.org/
```

### Specify File

Let us download another log file from the [RStudio CRAN](http://cran-logs.rstudio.com/) mirror and save it into a file using the `-o` option. 

```{bash c66a, eval=FALSE}
curl http://cran-logs.rstudio.com/2019/2019-09-08.csv.gz -o sept_08.csv.gz 
```

```{bash c66b, echo=FALSE, eval=FALSE}
cd cline
curl http://cran-logs.rstudio.com/2019/2019-09-08.csv.gz -o sept_08.csv.gz 
```

Another way to save a downloaded file is to use `>` followed by the name of the file as shown in the below example.

```{bash c89a, eval=FALSE}
curl http://cran-logs.rstudio.com/2019/2019-09-01.csv.gz > sep_01.csv.gz
```

```{bash c89b, echo=FALSE, eval=FALSE}
cd cline
curl http://cran-logs.rstudio.com/2019/2019-09-01.csv.gz > sep_01.csv.gz
```

### Download Silently

The `-s` option will allow you to download files silently. It will mute `curl` and will not display progress meter or error messages.

```{bash c67a, eval=FALSE}
curl http://cran-logs.rstudio.com/2019/2019-09-01.csv.gz -o sept_01.csv.gz -s
```

```{bash c67b, echo=FALSE, eval=FALSE}
cd cline
curl http://cran-logs.rstudio.com/2019/2019-09-01.csv.gz -o sept_01.csv.gz -s
ls
```

## R Functions

In R, we can use `download.file()` to download files from the internet. The following packages offer functionalities that you will find useful.

- [curl](https://cran.r-project.org/package=curl)
- [R.utils](https://cran.r-project.org/package=R.utils)
- [pingr](https://cran.r-project.org/package=pingr)

```{r r_data_transfer, echo=FALSE}
cname   <- c("`wget`", "`curl`", "`hostname`", "`ping`", "`nslookup`")
descrip <- c("`download.file()`", 
             "`curl::curl_download()`",
             "`R.utils::getHostname.System()`",
             "`pingr::ping()`",
             "`curl::nslookup()`")
data.frame(Command = cname, R = descrip) %>% 
  kable() %>% 
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "responsive")
  )
```