# File Compression {#compress}

In this chapter, we will explore file compression using the following

- `tar`
- `gzip`
- `zip`
- `unzip`

## tar

The `tar` command is used for file compression. It works with both `tar` and
`tar.gz` extensions. It is used to

- list files 
- extract files 
- create archives
- append file to existing archives

`tar` creates, maintains, modifies, and extracts files that are archived in the tar format. Tar stands for **t**ape **ar**chive and is an archiving file format.

```{r table_tar, echo=FALSE}
cname   <- c("`tar tvf`", "`tar tvfz`", "`tar xvf`", "`tar xvfz`", "`tar cvf`", 
             "`tar cvfz`", "`tar rvf`", "`tar rvfz`")
descrip <- c("List an archive",
             "List a gzipped archive",
             "Extract an archive", 
             "Extract a gzipped archive",
             "Create an uncompressed tar archive",
             "Create a tar gzipped archive", 
             "Add a file to an existing archive", 
             "Add a file to an existing gzipped archive")
data.frame(Command = cname, Description = descrip) %>% 
  kable() %>% 
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "responsive")
  )
```

We will use different options along with the `tar` command for listing, extracting, creating and adding files. The `vf` (**v** stands for verbosely show .tar file progress and **f** stands for file name type of the archive file) option is common for all the above operations while the following are specific.

- `t` for listing
- `x` for extracting
- `c` for creating
- `r` for adding files

While dealing with `tar.gz` archives we will use `z` in addition to `vf` and the above options. 

### List

Let us list all the files & folders in `release_names.tar`. As mentioned above. to list the files in the archive, we use the `t` option.

```{bash c70a, eval=FALSE}
tar -tvf release_names.tar 
```

```{bash c70b, echo=FALSE}
cd cline
tar -tvf release_names.tar 
```

### Extract

Let us extract files from `release_names.tar` using the `x` option in addition to `vf`.

```{bash c71a, eval=FALSE}
tar -xvf release_names.tar 
ls
```

```{bash c71b, echo=FALSE}
cd cline
rm release_names.txt release_names_18.txt release_names_19.txt
tar -xvf release_names.tar 
ls
```

### Add 

To add a file to an existing archive, use the `r` option. Let us add `release_names_18.txt` and `release_names_19.txt` to the archive we created in the previous step. 

```{bash c69a, eval=FALSE}
tar -rvf release_names.tar release_names_18.txt release_names_19.txt
```

```{bash c69b, echo=FALSE}
cd cline
tar -rvf release_names.tar release_names_18.txt release_names_19.txt
```

### Create 

Using the `c` option we can create tar archives. In the below example, we are using a single file but you can specify multiple files and folders as well.

```{bash c68a, eval=FALSE}
tar -cvf pkg_names.tar pkg_names.txt
```

```{bash c68b, echo=FALSE}
cd cline
tar -cvf pkg_names.tar pkg_names.txt
```

## gzip

```{r table_gzip, echo=FALSE}
cname   <- c("`gzip`", "`gzip -d`", "`gzip -c`", "`zip -r`", "`zip`",                       "`unzip`", "`unzip -d`", "`unzip -l`")
descrip <- c("Compress a file",
             "Decompress a file",
             "Compress a file and specify the output file name",
             "Compress a directory",
             "Add files to an existing zip file",
             "Extract files from a zip files",
             "Extract files from a zip file and specify the output file name",
             "List contents of a zip file")
data.frame(Command = cname, Description = descrip) %>% 
  kable() %>% 
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "responsive")
  )
```


`gzip`, `gunzip`, and `zcat` commands are used to compress or expand files in the GNU GZIP format i.e. files with `.gz` extension

### Compress 

Let us compress `release_names.txt` file using `gzip`.

```{bash c72a, eval=FALSE}
gzip release_names.txt
ls
```

```{bash c72b, echo=FALSE}
cd cline
gzip release_names.txt
ls
```

```{bash c72c, echo=FALSE}
cd cline
gunzip release_names.txt.gz
```


### Decompress 

Use the `-d` option with `gzip` to decompress a file. In the below example, we decompress the `sept_15.csv.gz` file (downloaded using `wget` or `curl` earlier). You can also use `gunzip` for the same result.


```{bash c73a, eval=FALSE}
gzip -d sept_15.csv.gz
ls
```

```{bash c73b, echo=FALSE}
cd cline
gzip -d sept_15.csv.gz
ls
```

### Specify Filename

Use `-c` and `>` to specify a different file name while compressing using `gzip`. In the below example, `gzip` will create `releases.txt.gz` instead of `release_names.txt.gz`.

```{bash c90a, eval=FALSE}
gzip -c release_names.txt > releases.txt.gz
ls
```

```{bash c90b, echo=FALSE}
cd cline
gzip -c release_names.txt > releases.txt.gz
ls
```

## zip & unzip

`zip` creates ZIP archives while `unzip` lists and extracts compressed files in a ZIP archive.

### List 

Let us list all the files and folders in `main_project.zip()` using `unzip` and the `-l` option. 

```{bash c97a, eval=FALSE}
unzip -l main_project.zip
```

```{bash c97b, echo=FALSE}
cd cline
unzip -l main_project.zip
```

### Extract 

Using `unzip`, let us now extract files and folders from `zip_example.zip`.

```{bash c95a, eval=FALSE}
unzip zip_example.zip
```

```{bash c95b, echo=FALSE}
cd cline
unzip zip_example.zip
```

Using the `-d` option, we can extract the contents of `zip_example.zip` to a specific folder. In the below example, we extract it to a new folder `examples`.

```{bash c96a, eval=FALSE}
unzip zip_example.zip –d myexamples
```

```{r unzip_d, echo=FALSE}
system2("unzip", c("zip_example.zip", "-d", "examples"), TRUE)
```

### Compress 

Use the `-r` option along with `zip` to create a ZIP archive. In the below example, we create a ZIP archive of `myproject` folder.

```{bash c91a, eval=FALSE}
zip -r myproject.zip myproject
ls
```

```{bash c91b, echo=FALSE}
cd cline
zip -r myproject.zip myproject
```

We can compress multiple directories using `zip`. The names of the directories must be separated by a space as shown in the below example where we compress `myproject` and `mypackage` into a single ZIP archive.

```{bash c92a, eval=FALSE}
zip -r packproj.zip myproject mypackage
ls
```

```{bash c92b, echo=FALSE}
cd cline
zip -r packproj.zip myproject mypackage
```

### Add 

To add a new file/folder to an existing archive, specify the name of the archive followed by the name of the file or the folder. In the below example, we add the `bash.sh` file to the `myproject.zip` archive created in a previous step.

```{bash c93a, eval=FALSE}
zip myproject.zip bash.sh
```

```{bash c93b, echo=FALSE}
cd cline
zip myproject.zip bash.sh
```

## R Functions

### tar & tar.gz

In R, we can use the `tar()` and `untar()` functions from the `utils` package to handle `.tar` and `.tar.gz` archives.

```{r r_tar, echo=FALSE}
cname   <- c("`tar tvf`", "`tar tvfz`", "`tar xvf`", "`tar xvfz`", "`tar cvf`", 
             "`tar cvfz`")
descrip <- c("`utils::untar('archive.tar', list = TRUE)`",
             "`utils::untar('archive.tar.gz', list = TRUE)`",
             "`utils::untar('archive.tar')`",
             "`utils::untar('archive.tar.gz')`",
             "`utils::tar('archive.tar')`",
             "`utils::tar('archive.tar', compression = 'gzip')`")
data.frame(Command = cname, R = descrip) %>% 
  kable() %>% 
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "responsive")
  )
```

### zip & gzip

The [zip](https://cran.r-project.org/package=zip) package has the functionalities to handle ZIP archives. The `tar()` and `untar()` functions from the `utils` package can handle GZIP archives.

```{r r_gzip, echo=FALSE}
cname   <- c("`gzip`", "`gzip -d`", "`gzip -c`", "`zip -r`", "`zip`",                       "`unzip`", "`unzip -d`", "`unzip -l`")
descrip <- c("`utils::tar(compression = 'gzip'` / `R.utils::gzip()`",
             "`utils::untar()` / `R.utils::gunzip()`",
             "`utils::untar(exdir = filename)`",
             "`zip::zip()`",
             "`zip::zipr_append()`",
             "`zip::unzip()`",
             "`zip::unzip(exdir = dir_name)`",
             "`zip::zip_list()`")
data.frame(Command = cname, R = descrip) %>% 
  kable() %>% 
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "responsive")
  )
```