# System Info {#command-line-sysinfo}

In this chapter, we will explore commands that will allow us to

- display information about the system
- display memory usage information
- display file system disk space usage 
- exit the terminal
- run commands a superuser
- shutdown the system

```{r table_sys_mgmt, echo=FALSE}
cname   <- c("`uname`", "`free`", "`df`", "`exit`", "`sudo`", "`shutdown`")
descrip <- c("Display important information about the system",
             "Display free, used, swap memory in the system",
             "Display file system disk space usage",
             "Exit the terminal",
             "Run command as super user",
             "Shutdown the system")
data.frame(Command = cname, Description = descrip) %>% 
  kable() %>% 
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "responsive")
  )
```

## uname

The `uname` command is used to view important information about the system. Using additional options, we can

- print details about operating system
- hardware & software related information

```{r table_uname, echo=FALSE}
cname   <- c("`uname`", "`uname -mp`", "`uname -srv`", "`uname -n`", "`uname -a`")
descrip <- c("Print details about the current machine and the operating system running on it",
             "Hardware related information; machine & processor",
             "Software related information; operating system, release number and version",
             "Nodename of the system",
             "Print all available information system")
data.frame(Command = cname, Description = descrip) %>% 
  kable() %>% 
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "responsive")
  )
```

### Print all available information about the system

```{bash c104a}
uname -a
```

### Display free, used, swap memory in the system

```{bash c105a}
free
```

### Display file system disk space usage

```{bash c106a}
df
```

## Others

In this section, let us look at a few other useful commands that will allow us 
to

- see how long a command takes to execute
- delay activity
- display and clear command history list

```{r table_others, echo=FALSE}
cname   <- c("`time`", "`sleep`", "`sleep 1m`", "`sleep 1h`", "`history`", "`history -c`")
descrip <- c("See how long a command takes to execute",
             "Delay activity in seconds",
             "Delay activity in minutes",
             "Delay activity in hours",
             "Display command history list with line numbers",
             "Clear the command history list")
data.frame(Command = cname, Description = descrip) %>% 
  kable() %>% 
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "responsive")
  )
```

## Funny Commands

Below are a few funny commands for you to try out. Use `sudo apt-get install` to install fortune and banner before trying them.

```{r table_funny_commands, echo=FALSE}
cname   <- c("`fortune`", "`yes`", "`banner`", "`rev`")
descrip <- c("Poignant, inspirational & silly phrases",
             "Output a string repeatedly until killed",
             "ASCII banner", "Reverse each character")
data.frame(Command = cname, Description = descrip) %>% 
  kable() %>% 
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "responsive")
  )
```

## R Functions

In R, we can use `Sys.sleep()` to delay activity and `history()` to view command history.

```{r r_others, echo=FALSE}
cname   <- c("`sleep`", "`history`")
descrip <- c("`Sys.sleep()`",
             "`history()`")
data.frame(Command = cname, R = descrip) %>% 
  kable() %>% 
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "responsive")
  )
```
