---
title: "Navigating File System"
knitr:
  opts_knit:
    root.dir: "cline"
---

# Navigating File System {#sec-navigating-files-and-directories}

## Introduction

In this chapter, we will learn commands that will help us

- navigate between different folders/directories 
- return current working directory 
- list all the files & folders in a directory 
- create and delete directories

```{r table_nav_file_sys, echo=FALSE}
cname   <- c("`pwd`", "`ls`", "`cd`", "`mkdir`", "`rmdir`")
descrip <- c("Print working directory",
             "List directory contents",
             "Change current working directory", 
             "Create directory",
             "Remove/delete directory")
data.frame(Command = cname, Description = descrip) %>% 
  kable() %>% 
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "responsive")
  )
```

`pwd` displays the name of the **p**resent **w**orking **d**irectory. 

```{bash}
pwd
```

`ls` displays information about files and directories in the current directory along with their associated metadata such as 

- size
- ownership
- modification date 

With no options, it will list the files and directories in the current directory, sorted alphabetically.

```{bash}
ls
```

`cd` (**c**hange **d**irectory) changes the current working directory. It is among the most used commands as it allows the user to move around the file system.

```{bash}
cd r
pwd
```

`mkdir` will create new directory. It will allow you to set file mode (permissions associated with the directory) i.e. who can open/modify/delete the directory.

```{bash}
mkdir rfiles
ls
```

`rmdir` will remove empty directories from the file system. It can be used to remove multiple empty directories as well. If the directory is not empty, `rmdir` will not remove it and instead display a warning that the directory is not empty.

```{bash}
rmdir rfiles
ls
```

## Change Working Directory

Let us focus a bit more on changing working directory. The below table shows commands for changing working directory to

- up one level
- previous working directory
- home directory
- and root directory

```{r table_cd, echo=FALSE}
cname   <- c("`cd .`", "`cd ..`", "`cd -`", "`cd ~`", "`cd /`")
descrip <- c("Navigate into directory",
             "Go up one level",
             "Go to previous working directory", 
             "Change directory to home directory",
             "Change directory to root directory")
data.frame(Command = cname, Description = descrip) %>% 
  kable() %>% 
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "responsive")
  )
```

All files and directories stem from one main directory, the root directory. All the other directories in the system are sub-directories of the root directory and the root directory has no parent directory. It is represented by a single slash (`/`). `cd /` will change the current working directory to the root directory.

```{bash}
cd /
pwd
```

The parent directory i.e. the directory one level up from the current directory which contains the directory we are in now is represented by two dots (`..`). `cd ..` will change us into the parent directory of the current directory.

```{bash}
cd ..
pwd
```

The home directory is the directory we are placed in, by default, when we launch a new terminal session. It is represented by the tilde (`~`).

```{bash}
cd ~
pwd
```

To change into the previous working directory, use `cd -`.

```{bash}
cd -
pwd
```

The current working directory is represented by a single dot (`.`). `cd .` will change us into the current directory i.e. it will do nothing. 

## List Directory Contents

`ls` will list the contents of a directory. Using different arguments, we can

- list hidden files
- view file permissions, ownership, size & modification date
- sort by size & modification date

```{r table_ls, echo=FALSE}
cname   <- c("`ls`", "`ls -l`", "`ls -a`", "`ls -la`", "`ls -lh`", "`ls -lS`", "`ls -ltr`")
descrip <- c("List directory contents",
             "List files one per line",
             "List all files including hidden files", 
             "Display file permissions, ownership, size & modification date",
             "Long format list with size displayed in human readable format",
             "Long format list sorted by size",
             "Long format list sorted by modification date")
data.frame(Command = cname, Description = descrip) %>% 
  kable() %>% 
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "responsive")
  )
```

### List files one per line

```{bash}
ls -l
```

### List all files including hidden files

```{bash}
ls -a
```

### Display file permissions, ownership, size & modification date

```{bash}
ls -la
```

### Display size in human readable format

```{bash}
ls -lh
```

### Sort list by size

```{bash}
ls -lS
```

### Sort list by modification date

```{bash}
ls -ltr
```

## R Functions

In R, `getwd()` will return the current working directory. You can use `here()` from the [here](https://here.r-lib.org/) package as well. To change the current
working directory, use `setwd()`. The [fs](https://fs.r-lib.org/index.html) package provides useful functions for file operations. 

```{r r_nav_file_sys, echo=FALSE}
cname   <- c("`pwd`", "`ls`", "`cd`", "`mkdir`", "`rmdir`")
r_map   <- c("`getwd()` / `here::here()`",
             "`dir()` / `list.files()` / `list.dirs()` / `fs::dir_ls()` / `dir_info()`",
             "`setwd()`",
             "`dir.create()` / `fs::dir_create()`",
             "`fs::dir_delete()`")
data.frame(Command = cname, R  = r_map) %>% 
  kable() %>% 
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "responsive")
  )
```